---
title: "DGIdb data collection and processing"
output: html_document
date: "2024-10-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Input: Downloaded data files from DGIdb
### Output: Clean and processed data that can be used for training DGIdb

import libraries
```{r}
library(tidyverse)
library(mygene)
library(httr)
library(jsonlite)
```

# Step 1: read data
```{r}
# drug interaction data (contains genes and drugs they interact with)
# THIS IS THE DATA I WILL BE USING
dgidb_interactions <- read_tsv("../data/raw/interactions.tsv")

# genes and potential actions/characteristics they have ex. lipase, 
# cell surface, tumor suppressor etc
dgidb_categories <- read_tsv("../data/raw/categories.tsv")

# contains only drug information
dgidb_drugs <- read_tsv("../data/raw/drugs.tsv")

# contains only gene information
dgid_genes <- read_tsv("../data/raw/genes.tsv")
```

# Step 2: Explore data

```{r}
# CHECK FOR NAS
print(any(is.na(dgidb_interactions))) #there are NAs and NULLS

# check dimensions
print(dim(dgidb_interactions)) #98239 rows and 13 columns

#check columnames and their datatypes
print(colnames(dgidb_interactions))
print(str(dgidb_interactions))
```

```{r}
dgidb_interactions

# plot how many of these drugs are approved?
approved_drugs <- dgidb_interactions %>%
  group_by(approved) %>%
  summarize(count = n())

ggplot(approved_drugs, aes(x = approved, y = count)) + geom_bar(stat = "identity", fill = c("red", "gray", "forestgreen")) + labs(title = "Number of FDA approved drugs vs Not approved", x = "Approval Status", y = "Count") +
  theme_minimal() + geom_text(aes(label = count), vjust = -0.5) + scale_x_discrete(labels = c("FALSE" = "Not Approved", "NULL" = "NULL/NA", "TRUE" = "Approved"))

```



# Step 3: Map entrez gene ids to the gene numbers in original df
```{r}
# get rid of unnecessary columns and get rid of nas
new_dgidb_inter <- dgidb_interactions %>% filter(gene_name!= 'NULL') %>% dplyr::select(-gene_claim_name,-interaction_source_db_name:-interaction_type, -immunotherapy,-anti_neoplastic) %>% filter(!is.na(gene_name))

# split gene-concept-id into 2 columns to map entrez ids
new_dgidb_inter <- new_dgidb_inter %>% 
  separate(gene_concept_id, c("gene_id_type", "gene_num"), ":")

# map entrez to gene identifier types
# define function
map_id <- function(map_genes, geneid_type) {
  getGenes(map_genes, scopes = geneid_type,
           fields = c("entrezgene", "symbol"), species = "human")
}

# split up gene id types so entrez ids can be mapped to them
dgidb_hgnc_dat <- new_dgidb_inter %>% filter(gene_id_type == "hgnc")
get_hgnc_list <- as.list(dgidb_hgnc_dat$gene_num)

# get entrez mappings for hgnc genes
mapped_hgnc <- as_tibble(map_id(get_hgnc_list,"hgnc"))

# now do the same for ensemble ids
dgidb_ensemble_dat <- new_dgidb_inter %>% filter(gene_id_type == "ensembl")

get_ensembl_list <- as.list(dgidb_ensemble_dat$gene_num)
mapped_ensembl <- as_tibble(map_id(get_ensembl_list, "ensembl.gene"))

# now do the same for the ncbi
dgidb_ncbi_dat <- new_dgidb_inter %>% filter(gene_id_type=="ncbigene")
get_ncbigene_list <- as.list(dgidb_ncbi_dat$gene_num)

mapped_ncbigene <- as_tibble(map_id(get_ncbigene_list,"entrezgene"))

#create function that cleans the mygene mapping query
clean_mygene <- function(mygene_results){  
  final_mygene <- mygene_results %>% dplyr::rename(gene_num = query, mygene_symbol=symbol ) %>% filter(!is.na(entrezgene)) %>% dplyr::select(-X_id,-X_version,-notfound)
  
  return(final_mygene)
}

new_mygene_ncbi <- clean_mygene(mapped_ncbigene)
new_mygene_hgnc <- clean_mygene(mapped_hgnc)
new_mygene_ensembl <- mapped_ensembl %>% dplyr::rename(gene_num = query, mygene_symbol=symbol ) %>% filter(!is.na(entrezgene)) %>% dplyr::select(-X_id,-X_version)

new_mygene_ensembl
new_mygene_hgnc
new_mygene_ncbi
```


# Step 4: Join the entrez gene ids to the original dfs and get rid of extra columns
```{r}
entrez_ensemble <- dgidb_ensemble_dat %>% inner_join(new_mygene_ensembl, by = "gene_num") %>% dplyr::select(-gene_id_type:-gene_name) %>% 
  relocate(c("entrezgene","mygene_symbol"), .before ="interaction_score") %>% 
  relocate(c("drug_claim_name", "drug_concept_id", "drug_name"), .after = "mygene_symbol")

# check to see if gene name and mygene symbol match. 
print(identical(entrez_ensemble$gene_name, entrez_ensemble$mygene_symbol))

entrez_hgnc <- dgidb_hgnc_dat %>% inner_join(new_mygene_hgnc, by = "gene_num") %>% distinct() %>% dplyr::select(-gene_id_type:-gene_name) %>% 
  relocate(c("entrezgene","mygene_symbol"), .before ="interaction_score") %>% 
  relocate(c("drug_claim_name", "drug_concept_id", "drug_name"), .after = "mygene_symbol")

# did not match here. Gene names in original data do not match the ones caputured in the mygene query. Will be sticking to mygene query.
print(identical(entrez_hgnc$gene_name, entrez_hgnc$mygene_symbol)) 

entrez_ncbi <- dgidb_ncbi_dat %>% inner_join(new_mygene_ncbi, by= "gene_num") %>% distinct() %>% dplyr::select(-gene_id_type:-gene_name) %>% 
  relocate(c("entrezgene","mygene_symbol"), .before ="interaction_score") %>% 
  relocate(c("drug_claim_name", "drug_concept_id", "drug_name"), .after = "mygene_symbol")

# did not match here. Gene names in original data do not match the ones caputured in the mygene query. Will be sticking to mygene query.
print(identical(entrez_ncbi$gene_name, entrez_ncbi$mygene_symbol)) 

# combine all these data frames into one

new_genemapped_drugdata <- rbind(entrez_ensemble, entrez_hgnc, entrez_ncbi)
new_genemapped_drugdata
```

# Step 5: Map Pubchem ids to the drug ids in original data
```{r}
#filter out each type of drug id into their own df

# get rid of any rows with NULL values and separate drug concept id into two columns
drugid_process <- new_genemapped_drugdata %>% filter(drug_concept_id!="NULL") %>% 
  separate(drug_concept_id, c("drug_id_type", "drug_num"), ":")
drugid_process

```

### Look at the different drug ids
```{r}
table(drugid_process$drug_id_type)
```

### Develop funciton that will map a common drug id to all the ids in data frame
```{r}


# Function to query UniChem for each drug ID
map_drug_id <- function(drug_id, src_id, target_id) {
  # Construct API URL
  url <- paste0("https://www.ebi.ac.uk/unichem/rest/src_compound_id/", drug_id, "/", src_id, "/", target_id)
  
  # Make the API request
  response <- GET(url)
  
  # Check for a successful response
  if (status_code(response) == 200) {
    # Parse JSON content
    data <- content(response, "text", encoding = "UTF-8")
    json_data <- fromJSON(data)
    
    # Extract the target ID, or return NA if not found
    if (length(json_data) > 0) {
      return(json_data$`src_compound_id`[1])
    } else {
      return(NA)
    }
  } else {
    warning("Failed to retrieve data for drug ID: ", drug_id)
    return(NA)
  }
}

```


### Filter df by drug id and map
```{r}
# drug_id, src_id, target_id
drugbank_df <- drugid_process %>% filter(drug_id_type == 'drugbank')
drugbank_ids <- as.list(drugbank_df$drug_num)
drug_to_chembl <-  sapply('DB04459',map_drug_id, 2, 1)

```






