---
title: "DGIdb data collection and processing"
output: html_document
date: "2024-10-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Input: Downloaded data files from DGIdb
### Output: Clean and processed data that can be used for training DGIdb

import libraries
```{r}
library(tidyverse)
library(mygene)
library(httr)
library(jsonlite)
```

# Step 1: read data
```{r}
# drug interaction data (contains genes and drugs they interact with)
# THIS IS THE DATA I WILL BE USING
dgidb_interactions <- read_tsv("../data/raw/interactions.tsv")

# genes and potential actions/characteristics they have ex. lipase, 
# cell surface, tumor suppressor etc
dgidb_categories <- read_tsv("../data/raw/categories.tsv")

# contains only drug information
dgidb_drugs <- read_tsv("../data/raw/drugs.tsv")

# contains only gene information
dgid_genes <- read_tsv("../data/raw/genes.tsv")

# remove uneeded data
rm(dgidb_categories, dgid_genes) # removes data
gc() #frees up memory
```

# Step 2: Explore data

```{r}
# CHECK FOR NAS
print(any(is.na(dgidb_interactions))) #there are NAs and NULLS

# check dimensions
print(dim(dgidb_interactions)) #98239 rows and 13 columns

#check columnames and their datatypes
print(colnames(dgidb_interactions))
print(str(dgidb_interactions))
```



```{r}
dgidb_interactions

# plot how many of these drugs are approved?
approved_drugs <- dgidb_interactions %>%
  group_by(approved) %>%
  summarize(count = n())

ggplot(approved_drugs, aes(x = approved, y = count)) + geom_bar(stat = "identity", fill = c("red", "gray", "forestgreen")) + labs(title = "Number of FDA approved drugs vs Not approved", x = "Approval Status", y = "Count") +
  theme_minimal() + geom_text(aes(label = count), vjust = -0.5) + scale_x_discrete(labels = c("FALSE" = "Not Approved", "NULL" = "NULL/NA", "TRUE" = "Approved"))

```



# Step 3: Map entrez gene ids to the gene numbers in original df
```{r}
# get rid of unnecessary columns and get rid of nas
new_dgidb_inter <- dgidb_interactions %>% filter(gene_name!= 'NULL') %>% dplyr::select(-gene_claim_name,-interaction_source_db_name:-interaction_type, -immunotherapy,-anti_neoplastic) %>% filter(!is.na(gene_name))

# split gene-concept-id into 2 columns to map entrez ids
new_dgidb_inter <- new_dgidb_inter %>% 
  separate(gene_concept_id, c("gene_id_type", "gene_num"), ":")

# map entrez to gene identifier types
# define function
map_id <- function(map_genes, geneid_type) {
  getGenes(map_genes, scopes = geneid_type,
           fields = c("entrezgene", "symbol"), species = "human")
}

# split up gene id types so entrez ids can be mapped to them
dgidb_hgnc_dat <- new_dgidb_inter %>% filter(gene_id_type == "hgnc")
get_hgnc_list <- as.list(dgidb_hgnc_dat$gene_num)

# get entrez mappings for hgnc genes
mapped_hgnc <- as_tibble(map_id(get_hgnc_list,"hgnc"))

# now do the same for ensemble ids
dgidb_ensemble_dat <- new_dgidb_inter %>% filter(gene_id_type == "ensembl")

get_ensembl_list <- as.list(dgidb_ensemble_dat$gene_num)
mapped_ensembl <- as_tibble(map_id(get_ensembl_list, "ensembl.gene"))

# now do the same for the ncbi
dgidb_ncbi_dat <- new_dgidb_inter %>% filter(gene_id_type=="ncbigene")
get_ncbigene_list <- as.list(dgidb_ncbi_dat$gene_num)

mapped_ncbigene <- as_tibble(map_id(get_ncbigene_list,"entrezgene"))

#create function that cleans the mygene mapping query
clean_mygene <- function(mygene_results){  
  final_mygene <- mygene_results %>% dplyr::rename(gene_num = query, mygene_symbol=symbol ) %>% filter(!is.na(entrezgene)) %>% dplyr::select(-X_id,-X_version,-notfound)
  
  return(final_mygene)
}

new_mygene_ncbi <- clean_mygene(mapped_ncbigene)
new_mygene_hgnc <- clean_mygene(mapped_hgnc)
new_mygene_ensembl <- mapped_ensembl %>% dplyr::rename(gene_num = query, mygene_symbol=symbol ) %>% filter(!is.na(entrezgene)) %>% dplyr::select(-X_id,-X_version)

new_mygene_ensembl
new_mygene_hgnc
new_mygene_ncbi
```


# Step 4: Join the entrez gene ids to the original dfs and get rid of extra columns
```{r}
entrez_ensemble <- dgidb_ensemble_dat %>% inner_join(new_mygene_ensembl, by = "gene_num") %>% dplyr::select(-gene_id_type:-gene_name) %>% 
  relocate(c("entrezgene","mygene_symbol"), .before ="interaction_score") %>% 
  relocate(c("drug_claim_name", "drug_concept_id", "drug_name"), .after = "mygene_symbol")

# check to see if gene name and mygene symbol match. 
print(identical(entrez_ensemble$gene_name, entrez_ensemble$mygene_symbol))

entrez_hgnc <- dgidb_hgnc_dat %>% inner_join(new_mygene_hgnc, by = "gene_num") %>% distinct(.keep_all = TRUE) %>% dplyr::select(-gene_id_type:-gene_name) %>% 
  relocate(c("entrezgene","mygene_symbol"), .before ="interaction_score") %>% 
  relocate(c("drug_claim_name", "drug_concept_id", "drug_name"), .after = "mygene_symbol")

# did not match here. Gene names in original data do not match the ones caputured in the mygene query. Will be sticking to mygene query.
print(identical(entrez_hgnc$gene_name, entrez_hgnc$mygene_symbol)) 

entrez_ncbi <- dgidb_ncbi_dat %>% inner_join(new_mygene_ncbi, by= "gene_num") %>% distinct() %>% dplyr::select(-gene_id_type:-gene_name) %>% 
  relocate(c("entrezgene","mygene_symbol"), .before ="interaction_score") %>% 
  relocate(c("drug_claim_name", "drug_concept_id", "drug_name"), .after = "mygene_symbol")

# did not match here. Gene names in original data do not match the ones caputured in the mygene query. Will be sticking to mygene query.
print(identical(entrez_ncbi$gene_name, entrez_ncbi$mygene_symbol)) 

# combine all these data frames into one

new_genemapped_drugdata <- rbind(entrez_ensemble, entrez_hgnc, entrez_ncbi)
new_genemapped_drugdata
```


```{r}
# clear up memory again by getting rid of unnecessary data
rm(entrez_ensemble, entrez_hgnc, entrez_ncbi, dgidb_ensemble_dat, dgidb_hgnc_dat,
   dgidb_ncbi_dat, get_ensembl_list, get_hgnc_list, get_ncbigene_list,
   mapped_ensemb, mapped_hgnc, mapped_ncbigene, new_mygene_ensembl, new_mygene_hgnc,
   new_mygene_ncbi)

gc()
```


# Step 5: Map Chembl ids to the drug ids in original data
```{r}
#filter out each type of drug id into their own df

# add info that shows where the drug info was obtained
drug_source_info <- dgidb_drugs %>% dplyr::select(concept_id, source_db_name) %>% 
  dplyr::rename(drug_concept_id = concept_id)
# get rid of any rows with NULL values and separate drug concept id into two columns
drugid_process <- new_genemapped_drugdata %>% inner_join(drug_source_info, by= "drug_concept_id") %>% 
  filter(drug_concept_id!="NULL") %>% separate(drug_concept_id, c("drug_id_type", "drug_num"), ":")
drugid_process



```

### Look at the different drug ids
```{r}
table(drugid_process$drug_id_type)
```

```{r}
table(drugid_process$source_db_name)
```



### Develop function that will map a common drug id to all the ids in data frame
```{r}
split_into_chunks <- function(vec, chunk_size) {
  split(vec, ceiling(seq_along(vec) / chunk_size))
}

# Function to query UniChem for each drug ID
map_drug_id <- function(drug_id,  source_id) {
  # Make the API request
  headers = c(
    accept = "application/json",
    `Content-Type` = "application/json"
  )
  
  data <- paste0(
  '{ "compound": "', drug_id, '", ',
  '"sourceID": ', source_id, ', ',
  '"type": "sourceID"}'
)

  
  res <- httr::POST(url = "https://www.ebi.ac.uk/unichem/api/v1/connectivity", httr::add_headers(.headers=headers), body = data)
  parsed_response <- content(res, as = "parsed", type = "application/json")
  
  # Extract ChEMBL IDs from the response
  chembl_ids <- unlist(lapply(parsed_response$sources, function(source) {
    if (source$longName[[1]] == "ChEMBL") {
      return(source$compoundId[[1]])
    }
    return(NULL)
  }))

  # Create a data frame with source ID and ChEMBL IDs
  if (length(chembl_ids) > 0) {
    df <- data.frame(
      source_id = rep(drug_id, length(chembl_ids)),
      target_id = chembl_ids,
      stringsAsFactors = FALSE
    )
  } else {
    df <- data.frame(
      source_id = drug_id,
      target_id = NA,
      stringsAsFactors = FALSE
    )
  }

  return(df)
}



```


### Filter df by drug id and map drug id
```{r}


# Function to process drug IDs in chunks of 50
process_drug_ids <- function(drug_ids, source_id, chunk_size = 20) {
  # Split drug IDs into chunks of 50
  drug_chunks <- split_into_chunks(drug_ids, chunk_size)
  total_chunks <- length(drug_chunks)

  # Initialize an empty data frame to collect results
  final_df <- data.frame()

  # Create a progress bar
  pb <- txtProgressBar(min = 0, max = total_chunks, style = 3)

  # Start timer
  start_time <- Sys.time()

  # Iterate over each chunk and process
  for (i in seq_along(drug_chunks)) {
    chunk <- drug_chunks[[i]]

    # Use lapply to call the map_drug_id function for each drug ID in the chunk
    results_list <- lapply(chunk, function(drug_id) map_drug_id(drug_id, source_id))

    # Combine the individual data frames into one
    chunk_df <- bind_rows(results_list)

    # Append to the final data frame
    final_df <- bind_rows(final_df, chunk_df)

    # Update the progress bar
    setTxtProgressBar(pb, i)
  }

  # Close the progress bar
  close(pb)

  # Calculate total elapsed time in seconds
  total_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
  
  # Convert to minutes and seconds
  minutes <- floor(total_time / 60)
  seconds <- round(total_time %% 60)
  
  # Display total elapsed time
  cat("\nTotal processing time:", minutes, "minutes and", seconds, "seconds\n")


  return(final_df)
}

# drug_id, src_id, target_id

#TEST with these ids#
drug_ids <- c('DB04459', 'DB01002', 'DB00316', 'DB00150', 'DB00201')


```

### Process drug bank drug ids
```{r}

# df <- drugid_process
# drug_type <- "drugbank"
# source_num <- "2"
# filename <- "rugbank_mapping.RData"
map_newdrug_id <- function(df, drug_type, source_num, filename){
  druginfo_df <- drugid_process %>% filter(drug_id_type == drug_type)
  drug_ids <- as.list(druginfo_df$drug_num)
  olddrug_to_newdrug_id <-  process_drug_ids(drug_ids, source_num)
  save(olddrug_to_newdrug_id, file = filename) 
}

```


```{r}
drugbank_df <- map_newdrug_id(drugid_process, "drugbank", "2", "../data/raw/drugbank_mapping.RData")
```



### Process iuphar
```{r}
iuphar_to_chembl <- map_newdrug_id(drugid_process, "iuphar.ligand", "4", "../data/raw/iuphar_mapping.RData")

```

```{r}
rxcui_to_chembl <- map_newdrug_id(drugid_process, "rxcui", "47", "../data/raw/iuphar_mapping.RData")
```














































































































































































































